---
title: "Data"
author: "Jay Lee"
date: "5/22/2017"
output: github_document
---

```{r include = F}
library(knitr)
library(openintro)
library(tidyverse)
library(forcats)
library(oilabs)
library(gapminder)
library(NHANES)
opts_chunk$set(message = F)
```


Below each code output are the variables explicitly used in the exercises.

## Introduction to Data

### High school and beyond survey

Mentioned as an example in videos, not used in analysis. From `openintro` package.

```{r}
data(hsb2)
colnames(hsb2)
```

### Emails

From `openintro` package.

```{r}
data(email50)
colnames(email50)
```

`number`, `num_char`, `exclaim_mess`, `spam`

### Gapminder

From `gapminder` package.

```{r}
data(gapminder)
colnames(gapminder)
```

### Berkeley grad school admissions

Transformed from `UCBAdmissions` in `datasets`.

```{r}
data(UCBAdmissions)
ucb_admit <- as.data.frame(UCBAdmissions)
ucb_admit <- transform(ucb_admit[rep(1:nrow(ucb_admit), ucb_admit$Freq),-4])
colnames(ucb_admit)
```

All variables used.

### County info

Mentioned as an example in videos, not used in analysis. From `openintro` package.

```{r}
data(county)
colnames(county)
```

`name`, `state`

### US Regions

Transformed from `state` in `datasets`.

```{r}
us_regions <- cbind(as.data.frame(state.name),
                    as.data.frame(state.region)) %>%
  transmute(state = state.name,
            region = fct_recode(state.region, "Midwest" = "North Central"))
colnames(us_regions)
```

All variables used.

### UT Austin course evals

From `oilabs` package.

```{r}
data(evals)
colnames(evals)
```

`score`, `rank`, `ethnicity`, `gender`, `language`, `age`, `cls_students`, `cls_level`, `cls_profs`, `cls_credits`, `bty_avg`, `pic_outfit`, `pic_color`

## Exploratory Data Analysis

### Comics dataset

`comics`, adapted from [FiveThirtyEight](https://github.com/fivethirtyeight/data/tree/master/comic-characters)'s data.

Full variable list: "name", "id", "align", "eye", "hair", "gender", "gsm", "alive", "appearances", "first_appear", "publisher"

Variables used: `align`, `id`, `gender`, 

### Cars dataset

`cars`, source unknown.

Full variable list: "name", "sports_car", "suv", "wagon", "minivan", "pickup", "all_wheel", "rear_wheel", "msrp", "dealer_cost", "eng_size", "ncyl", "horsepwr", "city_mpg", "hwy_mpg", "weight", "wheel_base", "length", "width"

Variables used: `weight`, `hwy_mpg`, `pickup`, `city_mpg`, `suv`, `ncyl`, `eng_size`, `horsepwr`, `msrp`, `width`, `rear_wheel`

### County public health

`life`, source unknown. Mentioned as an example in videos, not used in analysis.

Full variable list: "state", "county", "expectancy", "income"

Variables used: `expectancy`, `state`

### Gapminder

From `gapminder` package.

```{r}
data(gapminder)
colnames(gapminder)
```

`continent`, `year`, `lifeExp`, `pop`

### Full email data

From `openintro` package.

```{r}
data(email)
colnames(email)
```

`spam`, `num_char`, `exclaim_mess`, `image`, `attach`, `dollar`, `number`

## Correlation and Regression

### North Carolina birth info

From `openintro` package.

```{r}
data(births)
ncbirths <- births
colnames(ncbirths)
```

`weeks`, `weight`, `mAge`

### Mammals

From `openintro` package.

```{r}
data(mammals)
colnames(mammals)
```

`BodyWt`, `BrainWt`

### 2010 baseball

From `openintro` package.

```{r}
data(mlbBat10)
colnames(mlbBat10)
```

`OBP`, `SLG`, `SB`, `HR`, `AB`

### Body dimensions

From `openintro` package.

```{r}
data(bdims)
colnames(bdims)
```

`hgt`, `wgt`, `sex`

### UK smoking

From `openintro` package.

```{r}
data(smoking)
colnames(smoking)
```

`age`, `amtWeekdays`

### Anscombe regression data

Transformed from `anscombe` in `datasets` package

Full variable list: "id", "set", "x", "y"

Variables used: `set`, `x`, `y`

### Noise dataset

`noise`, randomly generated normal values

Full variable list: "x", "y", "z"

Variables used: `x`, `y`

### Galton data

Source unknown, sets of `Galton_men` and `Galton_women` with the same variables

Full variable list: "family", "father", "mother", "sex", "height", "nkids"

Variables used: `height`, `father`, `mother`

### Textbook prices

Mentioned as an example in videos, not used in analysis. From `openintro` package.

```{r}
data(textbooks)
colnames(textbooks)
```

`amazNew`, `uclaNew`

## Foundations of Inference

### NHANES

Mentioned as an example in videos, not used in analysis. From `NHANES` package.

```{r}
data(NHANES)
colnames(NHANES)
```

`Gender`, `HomeOwn`, `SleepHrsNight`, `SleepTrouble`, `HealthGen`

### Gender discrimination

`disc`, source unknown.

Full list of variables: "promote", "sex"

All variables used

### Opportunity cost

`opportunity`, source unknown.

Full list of variables: "decision", "group"

All variables used.

### Polling data

`all_polls`, source unknown.

Full list of variables: "poll", "vote"

All variables used.