---
title: "Benford's Law"
output:
  html_document:
    css: ../lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(knitr)
instructor <- FALSE
knitr::opts_chunk$set(echo = instructor, eval = instructor)
```

```{r note}
# Instructor version, shows solutions

# Start off linking this to the probability lectures. Talk about some ways in which prob appears in nature??? Idk here
# There's a similar pattern called Zipf's law that says in a language, the 2nd most commonly used word is used half as much as the most common, the 3rd most used 1/3 as much, the 4th 1/4 as much, etc.
```

## 2009 Iran Election

On June 12 2009, the Republic of Iran held an election where President Mahmoud Ahmadinejad sought re-election against three challengers. When it was announced that Ahmadinejad had won with 62% of the vote, there were widespread allegations of election fraud.
There are many methods, both quantitative and qualitative, to detect election fraud. In this lab we will explore just one proposed method.

## Benford's Law

Benford's law observes that in many naturally-occuring phenomena with a quantitative measurement, some patterns can be observed in the digits of that measurement. Specifically, the law states that the first digit of such observations, rather than being distributed uniformly between the numbers 1-9, takes a decreasing log distribution. The code and plotted distribution for this are below. Investments with interest, populations of cities, and election results have all been observed by different groups to follow Benford's law.

```{r benford_code, echo = T, eval = T}
benfords_p <- data.frame(1:9, log10(1 + 1/1:9))
```

```{r benford_plot, eval = T}
library(ggplot2)
colnames(benfords_p) <- c("digit", "prop")
ggplot(benfords_p, aes(x = digit, y = prop)) +
  geom_bar(stat = "identity")
```

One purported method of detecting election fraud, then, is seeing if the first digit of the results for each reporting region fits the distribution specified by Benford's law. Try this out with the Iran data.

## Getting Started

### Load packages

For this lab we will read in the data using the `readr` package, explore it using the `dplyr` package and visualize it using the `ggplot2` package for data visualization. The data can be found in the `oilabs` package.

```{r load-packages, message = F}
library(readr)
library(dplyr)
library(ggplot2)
```

### Data

The data on Iran was collected from ____. The dataset contains 366 observations of 9 variables, where every row is a city. The first two variables identify the province and city the vote totals come from, the next four give vote totals for the four candidates (Mahmoud Ahmadinejad, Mohsen Rezaee, Mehdi Karroubi, and Mir-Hossein Mousavi), and the final three give total and procedural vote totals. 

```{r load_iran, message = F}
iran <- read_csv('iran.csv')
data(iran)
```

## Elections

### Iran, 2009

1.  Add a column to the `iran` dataset with the first digit of the votes cast for the winner, Mahmoud Ahmadinejad. Show the dataset now has this column.


Hint: To extract a subset of a string by index (position), try the `stringr::str_sub` command. For example, to pull "FIRST" from "FIRST TO LAST" we would run:

```{r hint, echo = T}
library(stringr)
str_sub(string = "FIRST TO LAST", start = 1, end = 5)
```

```{r solution1}
library(stringr)
iran <- iran %>%
  mutate(digit = as.numeric(str_sub(as.character(ahmadinejad), 1, 1)))
str(iran)
```

2.  Create a bar plot like the one above, showing the frequencies of first digits in the total vote count. Does this plot appear to match the ideal Benford's distribution?

```{r solution2}
ggplot(iran, aes(x = digit)) +
  geom_bar()
```

This result might be due to election fraud, but it might just be a result of a small sample size or other innocent factors. Try a similar method on data collected from the 2016 US presidential election.

***

```{r Mid-lab instructor notes}
# Break here to talk about OpenElections and go through an example (ND is presented here)
```

### United States, 2016

The [OpenElections](http://www.openelections.net) project is obtaining and standardizing precinct-level results from the 2016 US presidential election, among other US elections. To get the data, search OpenElections' [GitHub page](https://github.com/openelections/tracker/blob/master/state_status_2016.csv) and pick an election that is marked "CSV" or "baked raw".

3. Pick a state and use this data to create a plot of that state's first digit distribution by precinct. Use the number of votes cast for Hillary Clinton in each precinct.

```{r solution3}
# Example: ND
# Remind them to remove total counts

library(stringr)
ND <- read_csv("https://github.com/openelections/openelections-data-nd/raw/master/2016/20161108__nd__general__precinct.csv") %>%
  filter(Office == "President", Candidate == "CLINTON") %>%
  mutate(digit = as.numeric(str_sub(Votes, start = 1, end = 1)))

ggplot(ND, aes(x = digit)) +
  geom_bar()
```

4. Does the election you chose appear to fit the distribution better or worse than the Iran election?

```{r solution4}
# See plots at bottom of solution6 chunk
```

### Election Assessment

5. Develop a quantitative marker to describe how well this election fits the ideal Benford's distribution. Show how you created it.

**BONUS**: Normalize this measure (range from -1 to 1) so we can compare between states. Try using proportions.

```{r solution5}
# Example: some kind of chi-squared marker, created manually?
ND_count <- ND %>% count(digit) %>%
  mutate(nd_prop = n / sum(n)) %>%
  left_join(benfords_p, by = "digit") %>%
  mutate(chi_sq = (nd_prop - prop)^2 / prop)
chi_sq_ND <- nrow(ND_count) * sum(ND_count$chi_sq)
chi_sq_ND
```

6. Does this election fit the Benford's distribution better or worse than the Iran election, according to your marker?

```{r solution6}
iran_count <- iran %>% count(digit) %>%
  mutate(iran_prop = n / sum(n)) %>%
  left_join(benfords_p, by = "digit") %>%
  mutate(chi_sq = (iran_prop - prop)^2 / prop)
chi_sq_iran <- nrow(iran_count) * sum(iran_count$chi_sq)
chi_sq_iran

totals <- iran_count %>%
  left_join(ND_count, by = "digit") %>%
  select(digit, nd_prop, iran_prop, prop.x) %>%
  tidyr::gather(key = area, value = prop, -digit)

ggplot(totals, aes(x = digit, y = prop, fill = area)) +
  geom_bar(position = "dodge", stat = "identity")
ggplot(totals, aes(x = digit, y = prop, color = area)) +
  geom_density(stat = "identity")

```

